A dynamic programing algorithm to assess the reliability of inferences from network enrichment analysis

### 2018-05-02 doc/genesetdp.tex
Small edits

Gene enrichment analysis (GEA) is commonly used to infer function from sets of analytes such as genes, transcripts proteins or metabolites\cite{tavazoie1999systematic,khatri2012ten}. The technique predict which functional modules, such as complexes or pathways, that are overrepresentated among a set of identified analytes. One prominent application of the technique is expression analysis, where GEA is regularly used to assess alternation in pathway activity by examining significantly different concentrations of analytes between biological conditions, such as disease state or treatment group.

@@

Here, we wanted to determine a full score distribution, {\em i.e.} the number of ways, $N(s)$, we can pick $Q$ genes and obtaining a score, $s$. This allows us to evaluate our score against the null hyptesis, $H_0$ : ``The Q query genes are randomly selected from the genes in the genome''. This can be expressed as a mid $p$~value\cite{lancaster1961significance,hwang2001optimality} for obtaining a score $s$,

@@

\subsection*{Computation of the full score distribution}

@@

In our strive to calculate the score distribution, $N(s)$, we first want to investigate the number of ways, $N_a(s,c)$, to obtain a score $s$, when summing up links for $c$ genes with up to $\le a$ links to the pathway. This helps us formulate the recursion, using the fact that there are $k_a \choose b$ ways to select $b$ from $k_a$ elements, as
\begin{equation}
N_a(s,c)=\sum_{b=0}^{k_a}{k_a \choose b} N_{a-1}(s-ab,c-b),
\end{equation}
where $N_a(s',c')=0$ for all $s'\le 0$ or $c' \le -1$, with the exception for $N_{-1}(0,0)=1$.

@@

\begin{lstlisting}[caption={
The central part of the dynamic programing algorithm for finding $N(s)$. The function takes the vector of links per gene, $k_a$, as well as the number of query genes, $Q$ as an input. The function depends on two additional functions {\tt find\_maxscore(k,Q) }, which calculates the maximal score a query of size $Q$ can obtain, and {\tt ncr(a,b)}, which calculates $ a \choose b $.
}
, label=lst:gensetdp,captionpos=t,float,abovecaptionskip=-\medskipamount]
def genesetdp(k,Q):
    max_s = find_maxscore(k,Q)
    N = np.zeros((max_s+1,Q+1),dtype='u4')
    N[0,0] = 1

@@

We dowloaded network definitions and the BinoX software (on 2018/04/28) for comparisons from XXX \url{bitbucket/sonnhammerlab}. FIXME: add a more detailed description of the test set (i.e. organism and number of genes, pathways and network connections).

\section*{Results}

We implemented a Python program that read network and pathway definition files, and scored query sets according to Equation \ref{eq:sum_i}, the Dynamic Programming algorithm described in the Algorithm section, that enabled us to assign $p$~values according to Equation \ref{eq:pval}. We downloaded pathway and network definitions from the BinoX's website.

>>>>

Gene enrichment analysis (GEA) is commonly used to infer function from sets of analytes such as genes, transcripts proteins or metabolites\cite{tavazoie1999systematic,khatri2012ten}. The technique estimates which functional modules, such as complexes or pathways, that are overrepresentated among a set of identified analytes. One prominent application of the technique is expression analysis, where GEA is regularly used to assess alternation in pathway activity by examining significantly different concentrations of analytes between biological conditions, such as disease state or treatment group.

@@

Here, we wanted to determine a score distribution, {\em i.e.} the number of ways, $N(s)$, we can pick $Q$ genes and obtaining a score, $s$. This allows us to evaluate our score against the null hyptesis, $H_0$ : ``The Q query genes are randomly selected from the genes in the genome''. This can be expressed as a mid $p$~value\cite{lancaster1961significance,hwang2001optimality} for obtaining a score $s$,

@@

\subsection*{Computation of the score distribution}

@@

In our strive to calculate the score distribution, $N(s)$, we first want to investigate the number of ways, $N_a(s,c)$, to obtain a score $s$, when summing up links for $c$ genes with up to $\le a$ links to the pathway. This helps us formulate the recursion, using the fact that there are $k_a \choose b$ ways to select $b$ from $k_a$ elements, as
\begin{equation}
N_a(s,c)=\sum_{b=0}^{k_a}{k_a \choose b} N_{a-1}(s-ab,c-b),
\end{equation}
where $N_a(s,c)=0$ for all $s<0$, $c<0$ or $a<0$, with the exception for $N_{-1}(0,0)=1$.

@@

\begin{lstlisting}[caption={
The central part of the dynamic programing algorithm for finding $N(s)$. The function takes the vector of links per gene, $k_a$, as well as the number of query genes, $Q$ as an input. The function depends on two additional functions {\tt find\_maxscore(k,Q)}, which calculates the maximal score a query of size $Q$ can obtain, and {\tt ncr(a,b)}, which calculates $ a \choose b $.
}
, label=lst:gensetdp,captionpos=t,float,abovecaptionskip=-\medskipamount]
def genesetdp(k,Q):
    max_s = find_maxscore(k,Q)
    N = np.zeros((max_s+1,Q+1),dtype='float_')
    N[0,0] = 1

@@

We dowloaded network definitions and the BinoX software (on 2018/04/28) for comparisons from XXX \url{bitbucket/sonnhammerlab}. FIXME: add a more detailed description of the test set (i.e. organism and number of genes, pathways and network connections).
Troughout this study, BinoX was run with the default parameters.
\section*{Results}

We implemented a Python program that read network and pathway definition files, and scored query sets according to Equation \ref{eq:sum_i}, the Dynamic Programming algorithm described in the Algorithm section, that enabled us to assign $p$~values according to Equation \ref{eq:pval}. We downloaded pathway and network definitions from the BinoX's website and used the same network treshold as BinoX default ($0.7$).
